---
title: "Apply Type Functions"
author: "Lauren Cappiello"
date: "10/4/2022"
output: html_document
---

Loops are a great, relatively straightforward way to repeatedly execute a chunk of code. However, they aren't especially efficient. Enter: the `apply` type function:

- `replicate()`
- `apply()`
- `lapply()`
- `sapply()`
- `tapply()`
- `mapply()`

These functions run code chunks in a non-sequential way that is often more efficient than a loop (as long as the elements in your object aren't dependent on other elements in your object).

# `replicate()`

We will start with replicate because it's arguably the easiest of these functions to use. The `replicate` function repeats a function call `n` times.

```{r}
replicate(n=4, "Hello")

replicate(10, factorial(7))

# histogram of the means from 100 random samples of size n=10 from a standard normal distribution 
hist(replicate(100, mean(rnorm(10))),
     main = "", xlab="Means") 
```

### A Similar Function for Simple Value Replication

The function `rep` replicates the values in the first argument. This is *not* part of the `apply` family, but may serve a similar purpose to `replicate`.

Suppose I want to represent all of the possible combinations when rolling two four-sided dice. 

```{r}
v1 <- rep(1:4, times=4) # replicate the sequence 1:4, four times
v2 <- rep(1:4, each=4) # replicate 1:4, with each number replicated 4 times (in a row)
data.frame(v1, v2)
```

# `apply()`

The `apply` function applies a given function to the rows or columns of matrices (or arrays). It assembles the returned values into a vector, array, or list, which it returns. 

The `apply()` arguments:

- `X`: an array (matrix)
- `MARGIN` 
    - `1` indicates rows
    - '2' indicates columns
    - `c(1,2)` indicates rows and columns
- `FUN`: the function to be applied
- `...`: additional arguments to be passed to `FUN`

```{r}
data <- matrix(1:9, nrow=3, ncol=3)
# the following is equivalent to the command: colMeans(data)
apply(data, 2, mean) # data, columns, mean --> get column means

# the following is equivalent to the command: rowSums(data)
apply(data, 1, sum) # data, rows, sum --> get row sums
```

We can also use `apply` functions on user-defined functions.

```{r}
# Define the function within the apply statement:
apply(data, 2, function(x){
    y <- sum(x)^2 # sum of the input vector (here a column) squared
    return(y)
  }
)

# Define the function outside of the apply statement:
fn <- function(x){
    y <- sum(x)^2 # sum of the input vector (here a column) squared
    return(y)
}
apply(data, 2, fn)
```

The values that `apply()` returns depends on the function `FUN`. 

- If `FUN` returns an element of length 1, then `apply` will return a vector. 
- If `FUN` always returns an element of length $n>1$, then `apply` will return a matrix with n rows, and the number of columns will correspond to how many rows/columns were iterated over. 
- Finally, if `FUN` returns an object that would vary in length, then `apply` will return a list where each element corresponds to a row or column that was iterated over. 

In short, `apply` prioritizes returning a vector, array (matrix), and list (in that order). What is returned depends on the output of `FUN`.

Note: running `apply` on a data frame will cause R to convert the data frame using `as.matrix`. This is often not what we want, so be cautious doing that. 

### Example: Extra Arguments, Array Output

```{r}
x <- cbind(x1 = 3, x2 = c(4:1, 2:5))

fun1 <- function(x, c1, c2){
  mean_vec <- c(mean(x[c1]), mean(x[c2]))
  return(mean_vec)
}

apply(x, 1, fun1,  c1 = "x1", c2 = c("x1","x2"))
```

### Example: List Output

```{r}
mat <- matrix(c(-1, 1, 0, 
                2, -2, 20, 
                62,-2, -6), nrow = 3)

CheckPos <- function(Vec){
  # Subset values of Vec that are even
  PosVec <- Vec[Vec > 0]
  
  # Return only the even values
  return(PosVec)
}

# Check Positive values by column 
apply(mat, 2, CheckPos)
```

### On Your Own

Use an apply function to find the interquartile range (`IQR()`) of each variable in the `ChickWeight` data. (This dataset is built into R.)

```{r}

```

# `lapply()`

The `lapply` function is used to apply a function to each element of a list. It collects the returned values into another list, which it returns. 

Arguments:

- `X`: a list
- `FUN`: the function to be applied
- `...`: additional arguments to be passed to `FUN`

```{r}
data_lst <- list(item1 = 1:5,
                item2 = seq(4,36,8),
                item3 = c(1,3,5,7,9))
data_vector <- 1:8

lapply(data_lst, sum)
lapply(data_vector, sum) # lapply performs an `as.list` command on X if it's not already a list
```

```{r}
x <- list(a = 1:10, 
          beta = exp(-3:3), 
          logic = c(TRUE,FALSE,FALSE,TRUE))

# compute the list mean for each list element
lapply(x, mean)
```

Consider the built-in data set `iris`. If we use the `as.list()` function, each column is converted into an element of a list.
```{r}
head(iris)
str(as.list(iris))
```

So if we use `lapply()` in this case, it will iterate over the *columns*. We can find all values within a variable that are greater than the variable mean (for columns 1-4, the numeric variables).

```{r}
lapply(iris[,1:4], function(column){
  big_values <- column[column > mean(column)]
  return(big_values)
})
```

### On Your Own

Use `lapply` to find the range for each item in the list `data_lst` (which should already be in your R environment from an earlier code chunk).

# `sapply()`

The `sapply` function works basically the same as the `lapply` function. The primary difference is that `sapply` attempts to simplify the result into a vector or matrix (instead of a list). This simplification works the same way as in `apply`. 

```{r}
lapply(data_lst, sum) # returns a list

sapply(data_lst, sum) # returns a vector
```

### On Your Own

Use `sapply` to find the range for each item in the list `data_lst`.

# `tapply()`

The `tapply` function breaks the data set up into groups and applies a function to each group.

Arguments:

- `X`: A 1 dimensional object
- `INDEX`: A grouping factor or a list of factors
- `FUN`: The function to be applied

```{r}
data = data.frame(name=c("Amy","Jose","Ray","Kim","Sam","Eve","Bob"), 
                  age=c(24, 22, 21, 23, 20, 24, 21),
                  gender=factor(c("F","M","M","F","M","F","M"))) 
tapply(data$age, data$gender, min) # age, grouped by gender, min for each group
```

### On Your Own

For the `ChickWeight` data, use `tapply` find the mean weight for each chick. 

# `mapply()`

The `mapply` function is a multivariate version of `sapply`. It applies `FUN` to the first elements of each `...` argument, the second elements, the third elements, and so on.

Arguments:

- `FUN`: The function to be applied
- `...`: Arguments to vectorize over (vectors or lists of strictly positive length, or all of zero length).

```{r}
mapply(rep, times = 1:4, x = 4:1)
```
More information and examples: http://adv-r.had.co.nz/Functionals.html 