---
title: "ggplot2"
author: "Lauren Cappiello"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Graphics with ggplot2

The package `ggplot2` is used to produce graphics using a different system than base R plotting. 

```{r}
# install.packages(ggplot2)
library(ggplot2)
```

All such plots are composed of data and a mapping.

- Data: information you want to visualize
- Mapping: description of how the variables are mapped to aesthetic attributes.

There are five mapping components:

- **Layer**: a collect of geometric elements and statistical transformations.
    - Geometric elements are called **geoms**, they are what you see (points, lines, etc.)
    - Statistical transformations are called **stats**, they summarize the data (counting observations, making models)
- **Scales**: a map of the values in the data space to the aesthetic space.
    - Modifications of the color, shape, size
    - How to draw legends and axes
- Coordinate System: called **coord**, describe how data coordinates are mapped to the plane of the graphic.
    - Usually use Cartesian coordinate system.
- **Facet**: How to break up and display subsets of data as small multiples.
- **Theme**: control other points of display, like font size, and background color.

## Getting started

Every ggplot has three key components:

1. Data.
2. A set of **aes**thetic mappings between variables and visual properties.
3. At least one layer which describes how to render each observation, usually created with a **geom** function.

Example:

```{r}
ggplot(mpg,                         # Data
       aes(x = displ, y = hwy)) +   # aesthetic mapping
  geom_point()                      # geom layer 
```

## Adjusting Aesthetic Attributes

### Adjusting by variable

We can adjust aesthetic attributes by variable using the aes() function and the mapping argument.

Adjust the color for each point by `class`: 

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) + 
  geom_point()
```

Adjust the shape for each point by `drv`:

```{r}
ggplot(mpg, aes(x = displ, y = hwy, shape = drv)) + 
  geom_point()
```

Adjust the size of each point by `cyl`:

```{r}
ggplot(mpg, aes(x = displ, y = hwy, size = cyl)) + 
  geom_point()
```

### Adjusting for the entire plot

The plot changes based on if the color is assigned *outside* or *inside* the `aes()` function. 

This is because `aes()` is used to map the data to an aesthetic property of the map. Inside of `aes()` we let the aesthetic argument (`color`, `shape`, etc) equal a variable in the data set. The different values for that variable control what happens to that aesthetic. 

If we assign aesthetic value outside of the `aes()` function, we assign a value for the *entire data set*.

Change aesthetic attributes according to some variable:
```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(aes(color = class))
```

Attempt to change all points to blue:
```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(aes(color = "blue"))
```

Notice that this did not work because `color = "blue"` is *inside* `aes()`.

Change all points to blue:
```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(color = "blue")
```

## Faceting

*Faceting* creates tables of graphics by splitting the data into subsets and displaying the same graph for each subset. 

You can facet using `facet_wrap()` which is more flexible and automatically determines how to organize the data. Or, you can manually determine the facet structure using `facet_grid()`.

Facet wrap: let ggplot decide how many rows and columns to make.
```{r}
# break data up by class
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  facet_wrap(~class)
```

Facet grid: number of rows and columsn by number of unique values for each variable.
```{r}
# Row variable ~ Column variable. 
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  facet_grid(drv~class)
```

## More Examples

```{r}
# Example 1
ggplot(diamonds)+
  geom_boxplot(mapping=aes(y = price, x = cut, color = cut))
```

```{r}
# Example 2
ggplot(diamonds)+
  geom_point(mapping=aes(y = price, x = depth, size = carat, 
                         color = cut), pch = 1)+
  labs(title = "A Title is On This Plot")
```

```{r}
# Example 3
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill=color), 
           position = "fill")
```

```{r}
# Example 4
ggplot(diamonds, aes(x = clarity, group = cut, fill = cut)) +
  geom_bar(position = "dodge")
```

## Resources

- https://ggplot2-book.org/index.html
- https://www.stat.auckland.ac.nz/~ihaka/downloads/Interface98.pdf
- https://hbctraining.github.io/Intro-to-R-flipped/cheatsheets/data-visualization-2.1.pdf

## On Your Own

The code below generates an initial infection and plot for the SIR model. Use `ggplot2` to recreate the plot. (You may need to dig through those resources and do some googling.)

```{r}
set.seed(0)
nr <- 50
nc <- 50
p0 <- 0.1
sir.init <- matrix(rbinom(nr*nc, 1, p0), nr, nc)
image(sir.init, col = c("white","red"))
```