---
title: "Tidyverse"
author: "Lauren Cappiello"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# install.packages('tidyverse')
library(tidyverse)
```

The tidyverse is a collection of R packages including `ggplot2`, `dplyr`, `tidyr`, `readr`, `purr`, `tibble`, and `stringr`. There are approximately 20 of these in total.

The tidyverse philosophy is structured around "action" instead of "objects".

Tidyverse is designed for and often used for data science applications. It tends to be more beginner-friendly than base R, *but* (1) it can be a major deviation from base R and (2) it's not always flexible enough to do everything we want it to do. 

## Piping Operator

Tidyverse syntax is designed to be used with a "piping" operator (loaded in when we call the `tidyverse` library). 

The piping operator, `%>%` "feeds" things from left to right. 

```{r}
vec <- 1:10
vec %>% mean() # "feed" vec into the mean function
```

This is useful when we have a sequence of multiple operations and want to pipe the output of one into the next. 

### On Your Own

Convert the following base R code into a tidyverse approach using the piping operator:

```{r}
colMeans(subset(mtcars, vs = 1))
```

## Tibbles

A **tibble** is tidyverse's answer to the data frame. You may have seen things stored in the R environment as tibbles before - they're fairly widely used and function similarly to data frames. 

Unlike data frames, tibbles are designed to print in a way that doesn't fill your markdown document with 47 pages of output when you forget to comment out that `print` statement before compiling. (Similar to the `head` function in base R.)

```{r}
class(diamonds)
dim(diamonds)
diamonds
```

## Key Functions

Tidyverse is a gigantic collections of functions and objects, but these are a few of the main ones to help you get started.

Note, in general, for tidyverse help files arguments typically start with a ".", this in contrast to many of the base R help files where arguments are in all caps.

- `select()`: Select variables in a data frame.
- `filter()`: Subset a data frame, retaining all rows that satisfy your conditions.
- `arrange()`: Orders the rows of a data frame by the values of selected columns.
- `rename()`: Changes the names of individual variables using new_name = old_name syntax
- `mutate()`: Adds new variables and preserves existing ones.
- `group_by()`: Takes an existing tibble and converts it into a grouped tibble where operations can then be performed “by group”.
- `summarize()`/`summarise()`: Summarizes results for each group (rows), and summary statistics (columns).

Generally, the functions above have the following properties:

- The first argument is a data frame or a tibble.
- The subsequent arguments are used to determine what to do with the data-frame/tibble in the first argument.
- The returned value is a data frame or a tibble.
- The inputted data-frames/tibbles should be well formatted to start off with. Each row should be an observation, and each column should be a variable.
- When we refer to column names for the data frame or tibble in the first argument we do not need to use quotes around the column names.

### `select()`

Isolate particular columns:

```{r}
diamonds %>% select(price, cut)
```

We store this output as before:

```{r}
priceCut <- diamonds %>% select(price, cut)
priceCut
```

You can also use the operator “:”, and negative signs with the select() function. With the “name1:name2” operator we can select all columns between the column named “name1” and “name2”. With negative signs we can omit all variables that are preceded with a negative sign. These methods are typically not allowed in standard base R indexing when using names:

```{r}
# Select all columns between (and including) cut and price:
diamonds %>% select(price:cut)
```

```{r}
# Select all but price and cut
diamonds %>% select(-price, -cut)
```

### `filter()`

Isolate particular rows:

```{r}
mean(diamonds$depth) # mean of depth variable

diamonds %>% filter(depth > mean(depth)) # all rows where depth > mean(depth)
```

We can also filter on multiple conditions:

```{r}
diamonds %>% filter(depth > mean(depth), cut == "Good", price > 350)
```
### `arrange()`

The `arrange()` function works like `sort()` or `order()` in base R.

```{r}
diamonds %>% arrange(price)
```

### `rename()`

Allows us to rename variables (replaces `colnames()`).

Structure: `NewName = OldName`. If we want to rename the `cut` column to `Cut` (capitalized), this would look like:

```{r}
diamonds %>% rename(Cut = cut) #new = old
```

We can do this with as many columns at one time as we want.

### `mutate()`

The `mutate` function can be used to add or change a variable. 

```{r}
diamonds %>% mutate(price = price/100) # change price units from dollars to hundreds of dollars
```

We can change multiple columns at once and add new columns:

```{r}
diamonds %>% mutate(price = price/100, newVar = 10*depth + price)
```

The function `transmute()` is similar, but drops all other variables:

```{r}
diamonds %>% transmute(price = price/100, newVar = 10*depth + price)
```

### `group_by()`

This is often used with the `summarize` function to group sets of observations together.

### `summarize()` / `summarise()`

Group by `cut`, then calculate the mean `price` for each `cut`:

```{r}
diamonds %>% group_by(cut) %>% summarize(PriceMean = mean(price))
```

We can also do this with multiple groups and summary statistics.

```{r}
diamonds %>%
  group_by(cut, color) %>%
  summarize(PriceMean = mean(price), 
            PriceMedian = median(price))
```

### On Your Own

Convert the following line of code to code that uses piping operators and tidyverse functions.

```{r}
by(diamonds$price, diamonds$color, summary)
```

### On Your Own

Explain in plain English what each of the following code chunks does. You may need to examine the help files for some of the functions.

```{r}
diamonds %>%
  mutate(DxC = depth*carat) %>%
  group_by(cut) %>%
  summarise(AvgDxC = mean(DxC), 
            AvgCut = mean(price))
```

```{r}
diamonds %>%
  filter(cut == "Ideal") %>%
  select(cut, carat, depth, price)
```

```{r}
diamonds %>%
  filter(price >median(price)) %>%
  group_by(color) %>%
  summarize(mean_depth = mean(depth),
            min_depth = min(depth)) %>%
  arrange(mean_depth)%>%
  head(n = 10)
```