---
title: "Working With Data Sets"
author: "Lauren Cappiello"
date: "Lab 5"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Pulling Data Into R

Let's pull in the results for the survey we did on the first day of class. R will automatically create a data frame when we do this. You can pull in data directly from the web:

```{r}
stdat <- read.csv("https://lgpcappiello.github.io/teaching/stat128/SurveyResponses.csv")
```

...or you can use the import dataset wizard in RStudio. Data import can get a little complicated, so if the data are stored on your computer, this is what I recommend. You can find this wizard in the **Environment** tab. 

## Examining Real Data

Usually, we don't want to try to print out a lot of data right in the RMarkdown file or in the console. To avoid this, we can use the `View` command:

```{r}
View(stdat)
```

We can quickly examine a matrix or data frame using the `head` function, which shows only the first six rows:

```{r}
# head(stdat)
```