---
title: "If Statements"
author: "Lauren Cappiello"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## If

We can use `if` to execute commands only when certain conditions are met.

```{r eval=F}
if(<value>){
  # commands
}
```

Checks the condition, `<value>` and if it is true, runs the commands. (If `<value>` is false, R skips directly to the end of the curly brackets.)

```{r}
x <- 1
if(x > 0){
  print("positive")
}
```
Often `if` statements are set up like this with some relational operator as the `value`.

### On Your Own

Write an `if` statement that prints out a short phrase when x is equal to 1. 

## If Else

The operator `else` can be paired with an `if` statement. If the `if` statement is false, then the `else` code will run. If the `if` statement is true, the `if` code will run. (One or the other will run, but never both.)

```{r}
x <- -1

if(x > 0){ # check if x > 0
  print("x > 0") # runs if x > 0 is true
}else{ # note the placement of the else
  print("x <= 0")
}
```

### On Your Own

Copy and paste your code chunk from the last *On Your Own* prompt. Modify it to print out a different short phrase when x is not equal to 1.

## Else If

Sometimes we have reason to do a sequence of checks that are all related, where code will only run if all the previous statements were false and the current criteria is true. The operator `else if` can be used to implement this.

```{r}
x = -3

if(x > 0){
    type = "positive"
} else if (x < 0) {
    type = "negative"
} else if(x == 0){
    type = "zero"
} else {
    type = "Error"
}

type 
```

What's the logic here? 

Another example with `else if` used in a function:

```{r}
toyfun <- function(X,Y,Do){
  if(Do == "Add"){
    Z = X+Y

  }else if(Do =="Subtract"){
    Z = X-Y

  }else if(Do =="Multiply"){
    Z = X*Y

    
  }else if(Do =="Penguin"){
    Z = c("<('' )")
    
  } else{
    Z = c(X,Y)
  }
  
  return(Z)
}
toyfun(2,4,"Add")
toyfun(2,4,"Subtract")
toyfun(2,4,"Penguin")
toyfun(2,4,"typo")
```

### On Your Own

Implement a sequence that checks if a logical statement is `TRUE` or `FALSE`. Use the final `else` statement to allow for the possibility that there is some error and the input is neither true nor false. If `TRUE`, print "yes"; if `FALSE`, print "no"; otherwise print "error". 

## IfElse

Finally, we have a shortcut function that can sometimes help keep our code clean.

```{r eval=F}
ifelse(test, yes, no)
```

- `test` is the logical statement 
- `yes` is what the code does if `test` results in `TRUE`
- `no` is what the code does if `test` results in `FALSE`

```{r}
x <- 0
x.info <- ifelse(x < 0, "negative", "nonnegative"); x.info
```

We can also pass the `ifelse` function a vector and it will check each element individually. For example, in the mtcars data set, we can find the proportion of cars that have `mpg > 25` and `hp > 60`

```{r}
fast_efficient <- ifelse(mtcars$mpg > 25 & mtcars$hp>60, TRUE, FALSE)
sum(fast_efficient)/length(fast_efficient)
```

### On Your Own

Modify your code from the second *On Your Own* prompt to use an `ifelse` function. 

## Nested If Chains

```{r}
x <- 105
if(x>0){
  if(x>100){
     type = "large positive number"
  } else {
     type = "positive number"
  }
} else if(x<0) {
  type = "negative number"
} else if(x==0){
  type = "zero"
}else {
  type = "Error"
}

type 
```

We can also nest `ifelse` statements by making the `yes` and `no` results other `ifelse` statements, but this can get tough to follow pretty quickly! 

```{r}
x <- 0

numInf <- ifelse(x > 0, "positive", 
                 ifelse(x < 0, "negative", 
                        ifelse(x == 0, "zero", "error")))
numInf
```