---
title: "1.4 Frequency Distributions"
author: "Dr. Lauren Parker"
output:
  ioslides_presentation: 
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Categorical Variables

**Frequency** (**count**): the number of times a particular value occurs.

A **frequency distribution** lists each distinct value with its frequency. 

| Class | Frequency | 
| :-: | :-: |
| freshman | 12 |
| sophomore | 10 |
| junior | 3 |
| senior | 5 |

***

A **bar plot** is a graphical representation of a frequency distribution. Each bar's height is based on the frequency of the corresponding category. 

```{r}
df <- data.frame(Level=c("freshman","sophomore","junior","senior"),
                Count=c(12, 10, 3, 5))
df$Level <- ordered(df$Level, levels = c("freshman","sophomore","junior","senior"))
library(ggplot2)
library(scales)

ggplot(data=df, aes(x=Level, y=Count)) +
  geom_bar(stat="identity", fill="steelblue") +
  theme_minimal() + 
  ggtitle("Barplot of Class") + 
  scale_y_continuous(name="Frequency", breaks= pretty_breaks()) +
  theme(plot.title = element_text(hjust = 0.5))
```

## Relative Frequencies

**Relative frequency** is the ratio of the frequency to the total number of observations.

\[
  \text{relative frequency} = \frac{\text{frequency}}{\text{number of observations}}
\]

This is also called the **proportion**. 

The **percentage** can be obtained by multiplying the proportion by 100.

## Relative Frequency Distribution

A **relative frequency distribution** lists each distinct value with its relative frequency.

| Class | Frequency | Relative Frequency |
| :-: | :-: | :-: |
| freshman | 12 | $12/30 = 0.4$ |
| sophomore | 10 | $0.3333$ |
| junior | 3 | $0.1$ |
| senior | 5 | $0.1667$ |

## Numeric Variables

A **dot plot** shows a number line with dots drawn above the line. Each dot represents a single observation. 

```{r message=FALSE}
library(ggplot2)
set.seed(0)
dat <- data.frame(x=sample(1:10,50,TRUE))
ggplot(dat, aes(x = x)) +
  geom_dotplot(dotsize = 2, stackratio = 1, fill = "steelblue") + 
  scale_x_continuous(breaks = c(1:10)) +
  scale_y_continuous(NULL, breaks=NULL) + 
  labs(title = "Example Dot Plot", x = "Values", y = "") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

## Binning Complex Data

- We would also like to be able to visualize larger, more complex data sets. 
- This is hard to do using a dot plot! 
- Instead, we can do this using **bins**, which group numeric data into equal-width consecutive intervals.

## Example 

A random sample of weights (in lbs) from 12 cats:
 
\[\quad 6.2 \quad 11.6 \quad 7.2 \quad 17.1 \quad 15.1 \quad 8.4 \]
\[\quad	7.7 \quad	13.9 \quad	21.0 \quad	5.5 \quad 9.1 \quad	7.3 \]
 
- **Minimum** is 5.5 
- **Maximum** is 21

Lots of ways to break these into "bins", but what about...

- 5 - 10
- 10 - 15 
- 15 - 20
- 20 - 25

## Example

We've suggested bins

- 5 - 10
- 10 - 15 
- 15 - 20
- 20 - 25

Each has an equal width of 5 (that's good), but if we had a cat with a weight of exactly 15 lbs, would we use the second or third bin?? 

## Example

To make this clear, we need there to be no overlap. Instead, we could use:
 
| Weight |	Count |
| :-: | :-: |
| 5 - <10	| 7 |
| 10 - <15	| 2 |
| 15 - <20	| 2 |
| 20 - <25	| 1 |

Now, a cat with a weight of 15.0 lbs would be placed in the third bin (but not the second). 

## Example

We will visualize this using a **histogram**, which is a lot like a bar plot but for numeric data:

```{r}
wts <- c( 6.2, 11.6, 7.2, 17.1, 15.1, 8.4, 7.7, 13.9, 21.0, 5.5, 9.1, 7.3)
df <- data.frame(wts)
ggplot(data=df, aes(x=wts)) +
  geom_histogram(breaks=c(5,10,15,20,25), color="black", fill = "steelblue") +
  theme_minimal() + 
  ggtitle("Histogram of Cat Weights")+
  scale_y_continuous(name="Frequency") +
  scale_x_continuous(name="Weight (lbs)") +
  theme(plot.title = element_text(hjust = 0.5))
```

***

```{r}
wts <- c( 6.2, 11.6, 7.2, 17.1, 15.1, 8.4, 7.7, 13.9, 21.0, 5.5, 9.1, 7.3)
df <- data.frame(wts)
ggplot(data=df, aes(x=wts)) +
  geom_histogram(breaks=c(5,10,15,20,25), color="black", fill = "steelblue") +
  theme_minimal() + 
  ggtitle("Histogram of Cat Weights")+
  scale_y_continuous(name="Frequency") +
  scale_x_continuous(name="Weight (lbs)") +
  theme(plot.title = element_text(hjust = 0.5))
```

This is a **frequency histogram** because each bar height reflects the frequency of that bin. 

***

We can also create a **relative frequency histogram** which displays the relative frequency instead of the frequency:

```{r}
wts <- c( 6.2, 11.6, 7.2, 17.1, 15.1, 8.4, 7.7, 13.9, 21.0, 5.5, 9.1, 7.3)
df <- data.frame(wts)
ggplot(data=df, aes(x=wts)) +
  geom_histogram(aes(y = after_stat(count / sum(count))), 
                 breaks=c(5,10,15,20,25), color="black", fill = "steelblue") +
  theme_minimal() + 
  ggtitle("Histogram of Cat Weights")+
  scale_y_continuous(name="Relative Frequency") +
  scale_x_continuous(name="Weight (lbs)") +
  theme(plot.title = element_text(hjust = 0.5))
```

***

Notice that those last two histograms look the same *except for the numbers on the vertical axis*! 

- This gives us insight into the shape of the data **distribution**, literally how the values are distributed across the bins. 
- The part of the distribution that "trails off" to one or both sides is called a **tail** of the distribution. - When a histogram trails off to one side, we say it is **skewed**.

    - right-skewed if it trails off to the right
    - left-skewed if it trails off to the left
- Data sets with roughly equal tails are **symmetric**.

***

```{r}
par(mfrow=c(2,2))
set.seed(0)
x1 <- rpois(1000,2)
hist(x1, main="Right-Skewed Distribution", xlab="", ylab="", xaxt='n', yaxt='n')
x2 <- 8-x1
hist(x2, main="Left-Skewed Distribution", xlab="", ylab="", xaxt='n', yaxt='n')
x3 <- rnorm(1000)
hist(x3, main="Symmetric Distribution", xlab="", ylab="", xaxt='n', yaxt='n')
x4 <- c(rnorm(500), rnorm(500,5))
hist(x4, main="Symmetric Distribution", xlab="", ylab="", xaxt='n', yaxt='n')
```

***

We can also use a histogram to identify **modes**. 

```{r}
par(mfrow=c(2,2))
set.seed(0)
x3 <- rnorm(1000)
hist(x3, main="Unimodal", xlab="", ylab="", xaxt='n', yaxt='n')
x4 <- c(rnorm(500), rnorm(500,5))
hist(x4, main="Bimodal", xlab="", ylab="", xaxt='n', yaxt='n')
x1 <- c(rnorm(330), rnorm(330,5), rnorm(340,9))
hist(x1, main="Multimodal", xlab="", ylab="", xaxt='n', yaxt='n')
x2 <- c(rnorm(330), rnorm(330,5,0.5), rnorm(340,11,1.5))
hist(x2, main="Multimodal", xlab="", ylab="", xaxt='n', yaxt='n')
```

For numeric data, especially continuous variables, we think of modes as *prominent peaks*.

***

Finally, we can also "smooth out" these histograms and use a smooth curve to examine the shape of the distribution.

```{r}
par(mfrow=c(2,2))
set.seed(0)
x1 <- rpois(10000,2)
plot(density(x1, bw=0.6), main="Right-Skewed Distribution", xlab="", ylab="", xaxt='n', yaxt='n')
x2 <- 8-x1
plot(density(x2, bw=0.6), main="Left-Skewed Distribution", xlab="", ylab="", xaxt='n', yaxt='n')
x3 <- rnorm(10000)
plot(density(x3, bw=0.5), main="Symmetric Distribution", xlab="", ylab="", xaxt='n', yaxt='n')
x4 <- c(rnorm(5000), rnorm(5000,5))
plot(density(x4, bw=0.5), main="Symmetric Distribution", xlab="", ylab="", xaxt='n', yaxt='n')
```