---
title: "2.4 Position"
author: "Dr. Lauren Parker"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Measures of Position

The **interquartile range** (**IQR**) represents the middle 50% of the data. 

- This is another measure of variability!

## Percentiles

- Recall that the *median* cut the data in half: 50% of the data is below and 50% is above the median. 
- This is also called the **50th percentile**. 
- The **$p$th percentile** is the value for which $p$% of the data is below it.

***

To get the middle 50%, we will split the data into four parts:

| 1	| 2	| 3	| 4 |
| :-: | :-: | :-: | :-: |
| 25%	| 25%	| 25%	| 25% |

The 25th and 75th percentiles, along with the median, divide the data into four parts. 

## Quartiles

We call these three measurements the **quartiles**:

- **Q1**, the first quartile, is the median of the lower 50% of the data.
- **Q2**, the second quartile, is the median.
- **Q3**, the third quartile, is the median of the upper 50% of the data.

## Example

Consider {1, 2, 3, 4, 5, 6, 7, 8, 9, 10}

- Cutting the data in half: {1, 2, 3, 4, 5 | 6, 7, 8, 9, 10}
- So the median (Q2) is $\frac{5+6}{2}=5.5$.
- Q1 is the median of {1, 2, 3, 4, 5}, or 3
- Q3 is the median of {6, 7, 8, 9, 10}, or 8

**Note:** this is a "quick and dirty" way of finding quartiles. A computer will give a more exact result. 

## Interquartile Range

Then the interquartile range is
\[
  \text{IQR} = \text{Q3}-\text{Q1}
\]

- The IQR is resistant to extreme values. 

## Which measure to use?

- mean and standard deviation when the data are symmetric 
- median and IQR when the data are skewed

## Box Plots

Box plots summarize the data with 5 statistics plus extreme observations:

```{r}
set.seed(0)
x <- c(rpois(45,2),-2.5)
boxplot(x, ylim=c(-3,7), col="lightblue")
legend(x=0.62,y=6.8,legend="Upper Whisker",box.lty=0)
legend(x=0.68,y=3.8,legend="Q3",box.lty=0)
legend(x=1.2,y=2.8,legend="Median",box.lty=0)
legend(x=0.68,y=1.8,legend="Q1",box.lty=0)
legend(x=1.1,y=0.8,legend="Lower Whisker",box.lty=0)
legend(x=0.68,y=-1.7,legend="Potential Outlier",box.lty=0)
```

## Drawing a box plot:

1. Draw the vertical axis to include all possible values in the data.
2. Draw a horizontal line at the median, at $\text{Q1}$, and at $\text{Q3}$. Use these to form a box.
3. Draw the **whiskers**. The whiskers' upper limit is $\text{Q3}+1.5\times\text{IQR}$ and the lower limit is $\text{Q1}-1.5\times\text{IQR}$. The actual whiskers are then drawn *at the next closest data points within the limits*.
5. Any points outside the whisker limits are included as individual points. These are **potential outliers**. 

We won't draw box plots by hand, but understanding how they are drawn will help us understand how to interpret them!

## Outliers

(Potential) outliers can help us...

- examine skew (outliers in the negative direction suggest left skew; outliers in the positive direction suggest right skew).
- identify issues with data collection or entry, especially if the value of the outliers doesn't make sense.

## Descriptive Measures for Populations

- We've thought about calculating various descriptive statistics from a sample.
- Our long-term goal is to estimate descriptive information about a population. 
- At the population level, these values are called **parameters**. 

***

- When we find a measure of center, spread, or position, we use a sample to calculate a single value. 
- These single values are called **point estimates**.
    - They are used to *estimate* the corresponding population parameter. 

| Point Estimate | Parameter |
|-|-|
| sample mean: $\bar{x}$ | population mean: $\mu$ |
| sample standard deviation: $s$ | population standard deviation: $\sigma$ |


