---
title: "4.2 Discrete Random Variables"
author: "Dr. Lauren Parker"
output:
  ioslides_presentation: 
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Discrete Random Variables

A **random variable** is a quantitative variable whose values are based on chance. By "chance", we mean that you can't *know* the outcome before it occurs. 

A **discrete random variable** is a random variable whose possible values can be listed. 

## Notation

- Variables: $x$,$y$,$z$


- Random variables: $X$, $Y$, $Z$


- The event that the random variable $X$ equals $x$: $\{X=x\}$


- The probability that the random variable $X$ equals $x$: $P(X=x)$ 

## Probability Histograms

A **probability histogram** is a histogram where the heights of the bars correspond to the probability of each value. 

- For discrete random variables, each "bin" is one of the listed values. 

## Example: Probability Histograms

| Number of Siblings, $x$ | 0 | 1 | 2 | 3 | 4 |
| :- | :-: | :-: | :-: | :-: | :-: | 
| **Probability**, $P(X=x)$ | 0.200 | 0.425 | 0.275 | 0.075 | 0.025 | 

<br>
(Assume for the sake of the example that no one has more than 4 siblings.)

***
| Number of Siblings, $x$ | 0 | 1 | 2 | 3 | 4 |
| :- | :-: | :-: | :-: | :-: | :-: | 
| **Probability**, $P(X=x)$ | 0.200 | 0.425 | 0.275 | 0.075 | 0.025 | 

```{r}
barplot(c(0.2,0.425,0.275,0.075,0.025), space=0, col="lightblue",
        names.arg = c("0","1","2","3","4"),
        xlab="Number of Siblings",
        ylab = "Probability")
```

## The Mean of a Discrete Random Variable

The mean of a discrete random variable $X$ is denoted $\mu_X$. 

$$\mu_X = x_1P(X=x_1) + x_2P(X=x_2) + \dots + x_nP(X=x_n).$$

<br>

If it's clear which random variable we're talking about, we can just write $\mu$.

## Example

| Number of Siblings, $x$ | 0 | 1 | 2 | 3 | 4 |
| :- | :-: | :-: | :-: | :-: | :-: | 
| **Probability**, $P(X=x)$ | 0.200 | 0.425 | 0.275 | 0.075 | 0.025 | 

$$\mu = 0(0.200)+1(0.425)+2(0.275)+3(0.075)+4(0.025)=1.3$$
<br>
*Interpretation*: in a large number of independent observations of a random variable $X$, the mean of those observations will approximately equal $\mu$. 

## The Mean of a Discrete Random Variable

The mean of a random variable is also called the **expected value** or **expectation**. 

Since measures of center are meant to identify the most common or most likely, you can think of this as the value we *expect* to see (most often). 

## Law of Large Numbers

The larger the number of observations, the closer their average tends to be to $\mu$. This is known as the **law of large numbers**. 

## Example: Law of Large Numbers

Suppose I took a random sample of 10 people and asked how many siblings they have. $$2,2,2,2,1,0,3,1,2,0$$ In my random sample of 10, $\bar{x}=2$, which is a reasonable estimate but not that close to the true mean $\mu=1.3$. 

- A random sample of 30 gave me a mean of $\bar{x}=1.53$.
- A random sample of 100 gave me a mean of $\bar{x}=1.47$.
- A random sample of 1000 gave me a mean of $\bar{x}=1.307$.

## Standard Deviation of a Discrete Random Variable

\[ \sigma_X^2 = [x_1^2P(X=x_1) + x_2^2P(X=x_2) + \dots + x_n^2P(X=x_n)]-\mu_X^2\]

As before, the standard deviation is the square root of the variance: $$\sigma = \sqrt{\sigma^2}$$

## Example

Calculate the standard deviation of the Siblings variable. 

In general, a table is the best way to keep track of a variance calculation:
 
| $x$ | $P(X=x)$ | $xP(X=x)$ | $x^2$ | $x^2P(X=x)$ |
| :-: | :-: | :-: | :-: | :-: |
| 0 | 0.200 | 0 | 0 | 0 |
| 1 | 0.425 | 0.425 | 1 | 0.425 | 
| 2 | 0.275 | 0.550 | 4 | 1.100 | 
| 3 | 0.075 | 0.225 | 9 | 0.675 | 
| 4 | 0.025 | 0.100 | 16 | 0.400 | 
|  |  | $\mu$ = 1.3 |  | Total = 2.6 |

***

| $x$ | $P(X=x)$ | $xP(X=x)$ | $x^2$ | $x^2P(X=x)$ |
| :-: | :-: | :-: | :-: | :-: |
| 0 | 0.200 | 0 | 0 | 0 |
| 1 | 0.425 | 0.425 | 1 | 0.425 | 
| 2 | 0.275 | 0.550 | 4 | 1.100 | 
| 3 | 0.075 | 0.225 | 9 | 0.675 | 
| 4 | 0.025 | 0.100 | 16 | 0.400 | 
|  |  | $\mu$ = 1.3 |  | Total = 2.6 |

- The variance is $\sigma^2 = 2.6 - 1.3^2 = 0.9$
- The standard deviation is $\sigma = \sqrt{0.9} = 0.9539$