---
title: "5.2 Sampling Distributions"
author: "Dr. Lauren Parker"
output:
  ioslides_presentation: 
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Sampling Error

We want to use a sample to learn something about a population, but no sample is perfect! 

<br>

**Sampling error** is the error resulting from using a sample to estimate a population characteristic. 

## Sampling Error

If we use a sample mean $\bar{x}$ to estimate $\mu$, chances are that $\bar{x}\ne\mu$ (they might be close but... they might not be!). We will consider

- How close *is* $\bar{x}$ to $\mu$?
- What if we took many samples and calculated $\bar{x}$ many times? 
  - How would that relate to $\mu$? 
  - What would be the distribution of these values?

## Sampling Distribution

The distribution of a statistic (across all possible samples of size $n$) is called the **sampling distribution**. 

For a variable $x$ and given a sample size $n$, the distribution of $\bar{x}$ is called the **sampling distribution of the sample mean** or the **distribution of $\boldsymbol{\bar{x}}$**.

## Example

Suppose our population is the five starting players on a particular basketball team. We are interested in their heights (measures in inches). The full population data is

 | Player | A | B | C | D | E | 
 | :- | :-: | :-: | :-: | :-: | :-: | 
 | Height | 76 | 78 | 79 | 81 | 86 | 

The population mean is $\mu=80$. 

## Example

Consider all possible samples of size $n=2$:

 | Sample | A,B | A,C | A,D | A,E | B,C | B,D | B,E | C,D | C,E | D,E | 
 | :- | :-: | :-: | :-: | :-: | :-: |  :-: |  :-: |  :-: |  :-: |  :-: | 
 | $\bar{x}$ | 77 | 77.5 | 78.5 | 81.0 | 78.5 | 79.5 | 82.0 | 80.0 | 82.5 | 83.5 | 

There are 10 possible samples of size 2. 

- Of these samples, 10\% have means exactly equal to $\mu$.
    - For a *random* sample of size 2, you'd have a 10\% chance to find $\bar{x}=\mu$.
    
***

In general, the larger the sample size, the smaller the sampling error tends to be in estimating $\mu$ using $\bar{x}$. 

In practice, we have one sample and $\mu$ is unknown. 

## The sampling distribution of the sample mean

For the distribution of $\bar{X}$

- The mean of the distribution is $\mu_{\bar{X}}=\mu$.
- The standard deviation is $\sigma_{\bar{X}}=\sigma/\sqrt{n}$. 

We refer to the standard deviation of a sampling distribution as **standard error**. 

## Checkpoint

The mean living space for a detached single family home in the United States is 1742 ft$^2$ with a standard deviation of 568 square feet. For samples of 25 homes, determine the mean and standard error of $\bar{x}$. 

## The Distribution of $\bar{X}$

The plots show (A) a random sample of 1000 from a Normal(100, 25) distribution and (B) the approximate sampling distribution of $\bar{X}$ when X is Normal(100, 25).

```{r}
set.seed(0)
par(mfrow=c(1,2))
dat <- matrix(rnorm(1000000,100,25), nrow=1000, byrow=TRUE)
hist(dat[1,], freq=FALSE, main="Distribution of x", xlab="x", ylab="", yaxt="n")
hist(rowMeans(dat), freq=FALSE, main="Distribution of x-bar", xlab="x-bar", ylab="", yaxt="n")
```

## The Distribution of $\bar{X}$

In fact, if $X$ is Normal($\mu$, $\sigma$), then $\bar{X}$ is Normal($\mu_{\bar{X}}=\mu$, $\sigma_{\bar{X}}=\sigma/\sqrt{n}$). 

Surprisingly, we see a similar result for $\bar{X}$ even when $X$ is not normally distributed!

## The Central Limit Theorem

For relatively large sample sizes, the random variable $\bar{X}$ is approximately normally distributed *regardless of the distribution of* $X$: $$\bar{X}\text{ is Normal}(\mu_{\bar{X}}=\mu, \sigma_{\bar{X}}=\sigma/\sqrt{n}).$$

<br>

Notes

- This approximation improves with increasing sample size.
- In general, "relatively large" means sample sizes $n \ge 30$. 