---
title: "5.3 Developing Confidence Intervals"
author: "Dr. Lauren Parker"
output:
  ioslides_presentation: 
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

***

- A **point estimate** is a single-value estimate of a population parameter. 

- We say that a statistic is an **unbiased estimator** if the mean of its distribution is equal to the population parameter. 
    - Otherwise, it is a **biased estimator**. 

- Ideally, we want estimates that are unbiased with small standard error. 
    - For example, a sample mean (unbiased) with a large sample size (results in smaller standard error). 

*** 

Point estimates are useful, but they only give us so much information. The variability of an estimate is also important!

Take a look at these two boxplots:

```{r}
set.seed(0)
dat <- c(rnorm(100,0,0.5), runif(100,-5,5))
grp <- c(rep(1,100),rep(2,100))
boxplot(dat~grp, ylab="", xlab="")
```

***

```{r}
set.seed(0)
dat <- c(rnorm(100,0,0.5), runif(100,-5,5))
grp <- c(rep(1,100),rep(2,100))
boxplot(dat~grp, ylab="", xlab="")
```

- Both samples are size $n=100$ and have $\bar{x}=0$
- Variable 1 has a standard deviation of $\sigma=0.5$
- Variable 2 has standard deviation $\sigma=5$

## Confidence Intervals

A **confidence interval** is an interval of numbers based on the point estimate of the parameter (along with some other stuff). 

- Say we want to be 95\% confident about a statement. 
- In Statistics, this means that we have arrived at our statement using a method that will give us a correct statement 95\% of the time.

***

- Our best point estimate for $\mu$ (based on a random sample) is $\bar{x}$, so that value will make up the center of the interval. 
- To create an interval around $\bar{x}$, we will construct what is called the **margin of error**. 
    - We will use the variability of the data along with some normal distribution properties. 
    - This will look like $$z\times\frac{\sigma}{\sqrt{n}}$$ 
    - The value $z$ will come from the normal distribution and will be based on how confident we want to be, e.g., 95% confident.
    
***

Putting everything together, the 95% confidence interval is $$\left(\bar{x} - z_*\frac{\sigma}{\sqrt{n}}, \bar{x} + z_*\frac{\sigma}{\sqrt{n}}\right)$$ where $z_* = 1.96$. 

<br>

The value $1.96$ is chosen because $(-1.96 < Z < 1.96) = 0.95$ (this is what makes it a 95% confidence interval!).

## Interpreting a Confidence Interval

If an experiment is run infinitely many times, the true value of $\mu$ will be contained in 95\% of the intervals. 

```{r}
contains <- function(l,u,m){
  if(l < m & m < u) return(TRUE)
  else return(FALSE)
}
plot_ci <- function(lo, hi, m){
  par(mar=c(2, 1, 1, 1), mgp=c(2.7, 0.7, 0))
  k <- 100
  ci.max <- max(rowSums(matrix(c(-1*lo,hi),ncol=2)))

  xR <- m + ci.max*c(-1, 1)
  yR <- c(0, 41*k/40)

  plot(xR, yR, type='n', xlab='', ylab='', axes=FALSE)
  abline(v=m, lty=2, col='#00000088')
  axis(1, at=m, paste("mu = ",round(m,4)), cex.axis=1.15)
  #axis(2)
  for(i in 1:k){
    x <- mean(c(hi[i],lo[i]))
	  ci <- c(lo[i],hi[i])
	  if(contains(lo[i],hi[i],m)==FALSE){
		  col <- "#F05133"
		  points(x, i, cex=1.4, col=col)
#		  points(x, i, pch=20, cex=1.2, col=col)
		  lines(ci, rep(i, 2), col=col, lwd=5)
	  }
	  col <- 1
  	points(x, i, pch=20, cex=1, col=col)
	  lines(ci, rep(i, 2), col=col)
  }
}

set.seed(0)
population <- rnorm(1000000,80,25)
samp_mean <- rep(NA, 100)
samp_sd <- rep(NA, 100)
n <- 60
for(i in 1:100){
  samp <- sample(population, n) # obtain a sample of size n = 60 from the population
  samp_mean[i] <- mean(samp)    # save sample mean in ith element of samp_mean
  samp_sd[i] <- sd(samp)        # save sample sd in ith element of samp_sd
}
cv <- -qnorm(0.05/2)
lower_vector <- samp_mean - cv * samp_sd / sqrt(n) 
upper_vector <- samp_mean + cv * samp_sd / sqrt(n)
plot_ci(lower_vector, upper_vector, 80)
```

## Example 

The preferred keyboard height for typists is approximately normally distributed with $\sigma=2.0$. A sample of size $n=31$, resulted in a mean preferred keyboard height of $80 cm$. Find and interpret a 95\% confidence interval for keyboard height.

## Common mistakes

- It is NOT accurate to say that "the probability that $\mu$ is in the confidence interval is 0.95". 
    - The parameter $\mu$ is some fixed quantity and it's either in the interval or it isn't. 
- We are NOT "95\% confident that $\bar{x}$ is in the interval". 
    - The value $\bar{x}$ is some known quantity and it's always in the interval. 

## Checkpoint

Suppose I took a random sample of 50 Sac State students and asked about their SAT scores and found a mean score of 1112. Prior experience with SAT scores in the CSU system suggests that SAT scores are well-approximated by a normal distribution with standard deviation known to be 50. 
    
1. Find a 95\% confidence interval for Sac State SAT scores. 
2. Interpret your interval in the context of the problem. 
3. What is the width of your interval? If you want a narrower interval, what could you do? 