---
title: "5.4 Other Levels of Confidence"
author: "Dr. Lauren Parker"
output:
  ioslides_presentation: 
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

***

The 95\% confidence interval is common in research, but there's nothing inherently special about it. 

- You could calculate a 90\%, a 99\%, or even something like a 43.8\% confidence interval. 
- These numbers are called the **confidence level**.
    - They represent the proportion of times that the parameter will fall in the interval (if we took many samples).

## Confidence Intervals

The 100(1-$\alpha$)\% confidence interval for $\mu$ is given by $$\bar{x}\pm z_{\alpha/2}\frac{\sigma}{\sqrt{n}}$$ where $z_{\alpha/2}$ is the z-score associated with the $[1-(\alpha/2)]$th percentile of the standard normal distribution. 

## Critical Values

The value $z_{\alpha/2}$ is called the **critical value** ("c.v." on the plot, below). 

```{r}
x <- seq(-3,3,length.out=1000)
y <- dnorm(x)
x1 <- -3; x2 <- -qnorm(0.85); x3 <- qnorm(0.85); x4 <- 3
x.coord <- c(x[x > x1 & x < x2])
y.coord <- y[x > x1 & x < x2]
x.coord2 <- c(x[x > x3 & x < x4])
y.coord2 <- y[x > x3 & x < x4]
plot(x,y,type='l', xaxt='n', yaxt='n', ylab="",xlab="", ylim=c(-0.02,max(y)+0.01)) #
abline(h=0); abline(v=1100)
polygon(x=x.coord, y= c(0, y.coord[2:(length(y.coord)-1)], 0), col="mistyrose")
polygon(x=x.coord2, y= c(0, y.coord2[2:(length(y.coord)-1)], 0), col="mistyrose")
text(-1.6,0.03,"Area = a/2"); text(1.6,0.03,"Area = a/2"); text(0,0.1,"Area = 1-a")
text(x2, 0.-0.015, "-c.v."); text(x3, -0.015, "+c.v.")
```

## Common Critical Values

| Confidence Level | $\alpha$ | Critical Value, $z_{\alpha/2}$ |
| :-: | :-: | :-: |
| 90\% | 0.10 | 1.645 |
| 95\% | 0.05 | 1.96 | 
| 98\% | 0.02 | 2.326 |
| 99\% | 0.01 | 2.575 | 

## Checkpoint

Prior experience with SAT scores in the CSU system suggests that SAT scores are well-approximated by a normal distribution with standard deviation known to be 50. Suppose we have a random sample of 50 Sac State students with (sample) mean SAT score 1112. 

1. Calculate a 98\% confidence interval.
2. Calculate a 90\% confidence interval.
3. Interpret each interval in the context of the problem. Comment on how the intervals change as you change the confidence level.

## Breaking Down a Confidence Interval

$$\left(\bar{x}- z_{\alpha/2}\frac{\sigma}{\sqrt{n}}, \quad \bar{x} + z_{\alpha/2}\frac{\sigma}{\sqrt{n}}\right)$$ The key values are

- $\bar{x}$, the sample mean
- $\sigma$, the population standard deviation
- $n$, the sample size
- $z_{\alpha/2}$, the critical value $$P(Z > z_{\alpha/2}) = \frac{\alpha}{2}$$

## Breaking Down a Confidence Interval

$$\left(\bar{x}- z_{\alpha/2}\frac{\sigma}{\sqrt{n}}, \quad \bar{x} + z_{\alpha/2}\frac{\sigma}{\sqrt{n}}\right)$$

- The value of interest is $\mu$, the (unknown) population mean.
- The confidence interval gives us a reasonable range of values for $\mu$.

In addition, the formula includes 

- The standard error, $\frac{\sigma}{\sqrt{n}}$
- The margin of error, $z_{\alpha/2}\frac{\sigma}{\sqrt{n}}$

## Confidence Level

$$\left(\bar{x}- z_{\alpha/2}\frac{\sigma}{\sqrt{n}}, \quad \bar{x} + z_{\alpha/2}\frac{\sigma}{\sqrt{n}}\right)$$

If we can be 99\% confident (or even higher), why do we tend to "settle" for 95\%??

- What will higher levels of confidence do to this interval? 

## Sample Size

There is one other thing we can control in the confidence interval: the sample size $n$. 

- We can specify the confidence level and the maximum acceptable interval width and use these to determine sample size. 
- We know that $$\text{interval width} \ge 2z_{\alpha/2}\frac{\sigma}{\sqrt{n}}$$
- Letting interval width equal $w$, we can solve for $n$: $$ n \ge \left(2z_{\alpha/2}\frac{\sigma}{w}\right)^2$$ 

## Sample Size

Alternately, we may specify a maximum margin of error $m$ instead: $$ n \ge \left(z_{\alpha/2}\frac{\sigma}{m}\right)^2$$ 

- Once we've done a sample size calculation, we need a whole number for $n$. 
- Since $n \ge$ something, we will *always round up*.

## Checkpoint 

Suppose we want a 95\% confidence interval for the mean of a normally distributed population with standard deviation $\sigma=10$. It is important for our margin of error to be no more than 2. What sample size do we need?

## Some comments

- As desired width/margin of error decreases, $n$ will increase. 
- As $\sigma$ increases, $n$ will also increase. (More population variability will necessitate a larger sample size.)
- As confidence level increases, $n$ will also increase.

## Checkpoint

Prior experience with SAT scores in the CSU system suggests that SAT scores are well-approximated by a normal distribution with standard deviation known to be 50. 

Find the sample size required for a 98% confidence interval with maximum margin of error 10.