---
title: "Graphical Presentations"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

We would like to be able to get a quick overview of our data. One way to do this is with graphs.

A **bar plot** represents categorical data with bars. Each bar's height is based on the number of observations in that category.

```{r include=FALSE}
df <- data.frame(Level=c("freshman","sophomore","junior","senior"),
                Count=c(12, 10, 3, 5))
df$Level <- ordered(df$Level, levels = c("freshman","sophomore","junior","senior"))
library(ggplot2)
```

```{r}
p<-ggplot(data=df, aes(x=Level, y=Count)) +
  geom_bar(stat="identity", fill="steelblue")+
  theme_minimal()
p + ggtitle("Barplot of Class Level") + theme(plot.title = element_text(hjust = 0.5))
```

The bar plot above shows the class level breakdown for students in an Introductory Statistics course. Based on the plot, there are 5 seniors. How many sophomores are there? Freshman?

A **scatterplot** shows the relationship between two (numeric) variables.

```{r}
plot(mtcars$wt, mtcars$mpg, pch=16,
     main = "Scatterplot of Car Weight vs MPG",
     xlab="Weight (1000 lbs)", ylab="Miles Per Gallon")
```

At a glance, we can see that (in general) heavier cars have lower MPG.

Good data visualization is *easy to interpret*! Data visualization should have labels: title, axis labels, numbers/categories clearly marked, and a legend if necessary. 

We may also use graphs to examine the **distribution** of the data. A distribution relates values/categories to how often they occur. Think of distributing treats among a group of dogs. We could distribute the treats evenly or based on each dog's behavior/tricks.

A **dot plot** is a single-variable scatterplot.

```{r}
stripchart(mtcars$wt, method="stack",pch=16,xlab="title",
           main="Dot Plot")
```

We will see more graphical presentations as we move through this module. 

<center><b><a href="https://lgpcappiello.github.io/teaching/IntroStats/descriptives/plots">Next up: Graphical Displays</a> | <a href="https://lgpcappiello.github.io/teaching/IntroStats/descriptives/design.html">Back to: Experimental Design</a> | <a href="https://lgpcappiello.github.io/teaching/IntroStats">Return to Course Notes Homepage</a></center></b>
